### 2013-09-30 src2/expressive/abstract.tex
apa

\section*{Abstract}
Graphics Processing Units (GPUs) are powerful computing devices 
that with the advent of CUDA/OpenCL are becomming useful for general 
purpose computations. 
Obsidian is an embedded domain specific language that generates CUDA kernels
from functional descriptions.
A symbolic array construction allows us to guarantee that intermediate
arrays are fused away. However, the current array construction has
some drawbacks; in particular, arrays cannot be combined efficiently.
We add a new type of {\em push arrays} to the existing Obsidian system in 
order to solve this problem. The two array types complement each other,
and enable the definition of combinators that both
take apart and combine arrays, and that result in efficient generated code.
This extension to Obsidian is demonstrated on a sequence of
sorting kernels, with good results.
The case study also illustrates the use of combinators for expressing
the structure of parallel algorithms.
The work presented is preliminary, and 
the combinators presented must be generalised.
However, the raw speed of the generated kernels bodes well.
%\end{abstract}

>>>>

\subsection*{Abstract}
Graphics Processing Units (GPUs) are powerful computing devices 
that with the advent of CUDA/OpenCL are becomming useful for general 
purpose computations. 
Obsidian is an embedded domain specific language that generates CUDA kernels
from functional descriptions.
A symbolic array construction allows us to guarantee that intermediate
arrays are fused away. However, the current array construction has
some drawbacks; in particular, arrays cannot be combined efficiently.
We add a new type of {\em push arrays} to the existing Obsidian system in 
order to solve this problem. The two array types complement each other,
and enable the definition of combinators that both
take apart and combine arrays, and that result in efficient generated code.
This extension to Obsidian is demonstrated on a sequence of
sorting kernels, with good results.
The case study also illustrates the use of combinators for expressing
the structure of parallel algorithms.
The work presented is preliminary, and 
the combinators presented must be generalised.
However, the raw speed of the generated kernels bodes well.
%\end{abstract}

### 2013-09-30 src2/expressive/applications.tex
apa

\section{Application}
\label{sec:MARY}

\subsection{Sorting on a GPU}
In this section, we introduce combinators that express patterns
of computation on whole arrays, and show their application to the
development of fast sorting kernels.
By kernel, we mean a computation that is performed by multiple
threads, each performing the same computation, in a single block.
The computation is performed entirely on the GPU, operating on
a short array, which has been placed into shared memory.
On the GPU on which we perform measurements, the maximum number
of threads in a single block is 512. Thus, we will build and benchmark
a sequence of
kernels that sort 512 inputs.
Our first kernels are implemented using one thread per array element.
Next, we show how Push arrays allow us to move to having each
thread operate on two array elements, giving a substantial performance
improvement.

@@

\subsection{Describing Batcher's bitonic merger}

@@

\subsection{Modifying the bitonic merger}

@@

\subsection{New combinator implementations using Push arrays}

@@

\subsection{A sorter built from the balanced periodic merger}
First, we note that the balanced periodic merger contains multiple
uses of the now familiar vee-shaped pattern, see Figure~\ref{fig:periodicMerger}.
\begin{figure}
\centering
\includegraphics[scale=0.25]{./expressive/balanced}
\caption{16 input periodic balanced merging network}
\label{fig:periodicMerger}
\end{figure}

@@

\subsection{Measuring performance of the generated kernels}\label{sec:benchmarks}
\begin{figure}
\begin{codesize}
\begin{verbatim}
  unsigned int arrayLength = 1 << LOG_L_SIZE;
  unsigned int diff = LOG_L_SIZE - LOG_S_SIZE;
  unsigned int blocks = arrayLength / S_SIZE;
  unsigned int threads = S_SIZE / 2;

>>>>

\subsection{Application}
\label{sec:MARY}

\subsubsection{Sorting on a GPU}
In this section, we introduce combinators that express patterns
of computation on whole arrays, and show their application to the
development of fast sorting kernels.
By kernel, we mean a computation that is performed by multiple
threads, each performing the same computation, in a single block.
The computation is performed entirely on the GPU, operating on
a short array, which has been placed into shared memory.
On the GPU on which we perform measurements, the maximum number
of threads in a single block is 512. Thus, we will build and benchmark
a sequence of
kernels that sort 512 inputs.
Our first kernels are implemented using one thread per array element.
Next, we show how Push arrays allow us to move to having each
thread operate on two array elements, giving a substantial performance
improvement.

@@

\subsubsection{Describing Batcher's bitonic merger}

@@

\subsubsection{Modifying the bitonic merger}

@@

\subsubsection{New combinator implementations using Push arrays}

@@

\subsubsection{A sorter built from the balanced periodic merger}
First, we note that the balanced periodic merger contains multiple
uses of the now familiar vee-shaped pattern, see Figure~\ref{fig:periodicMerger}.
\begin{figure}
\centering
\includegraphics[scale=0.25]{./expressive/balanced}
\caption{16 input periodic balanced merging network}
\label{fig:periodicMerger}
\end{figure}

@@

\subsubsection{Measuring performance of the generated kernels}\label{sec:benchmarks}
\begin{figure}
\begin{codesize}
\begin{verbatim}
  unsigned int arrayLength = 1 << LOG_L_SIZE;
  unsigned int diff = LOG_L_SIZE - LOG_S_SIZE;
  unsigned int blocks = arrayLength / S_SIZE;
  unsigned int threads = S_SIZE / 2;

### 2013-09-30 src2/expressive/intro.tex
apa

\section{Introduction}

@@

\subsection{Embedded DSLs for GPGPU programming} 
We are aiming for a GPU programming language that is more
concise than mainstream
languages such as CUDA and OpenCL. 
Obsidian is a
domain specific embedded language (DSEL) implemented in Haskell.
When an Obsidian program is run, a representation of the program is created 
as a syntax tree. For more information on EDSL implementation see~\cite{COMPILEEDSL}.
The program representation generated when running an Obsidian program 
is compiled into CUDA code. We are also working on an OpenCL backend.

@@

%----------------------------------------------------------------------------
\subsection{Arrays in Obsidian}

@@

%----------------------------------------------------------------------------
\subsection{Sync and parallelism}

@@

\subsection{Drawbacks of Obsidian Arrays}
\label{sec:Drawbacks}

>>>>

\subsection{Introduction}

@@

\subsubsection{Embedded DSLs for GPGPU programming} 
We are aiming for a GPU programming language that is more
concise than mainstream
languages such as CUDA and OpenCL. 
Obsidian is a
domain specific embedded language (DSEL) implemented in Haskell.
When an Obsidian program is run, a representation of the program is created 
as a syntax tree. For more information on EDSL implementation see~\cite{COMPILEEDSL}.
The program representation generated when running an Obsidian program 
is compiled into CUDA code. We are also working on an OpenCL backend.

@@

%----------------------------------------------------------------------------
\subsubsection{Arrays in Obsidian}

@@

%----------------------------------------------------------------------------
\subsubsection{Sync and parallelism}

@@

\subsubsection{Drawbacks of Obsidian Arrays}
\label{sec:Drawbacks}

### 2013-09-30 src2/ifl/related.tex
apa

\section{Discussion} \label{sec:disc}
\subsection{Our influences}
As mentioned above, our earlier work on Lava has provided the inspiration for
the combinator-oriented or hardware-like style of programming that we are exploring in Obsidian.
On the other hand, the {\em implementation} of Obsidian has been much influenced by
{Pan}, an embedded language for image synthesis developed by 
Conal Elliot~\cite{PAN}. Because of the computational complexity of image generation, C 
code is generated. This C code can then be compiled by an optimising compiler. 
Many ideas from the paper 
``Compiling Embedded Languages'', describing the implementation of Pan have been 
used in the implementation of Obsidian~\cite{Elliott03:CompileDSEL-JFP}.

\subsection{Related work on GPU and GPGPU programming languages}
We cannot attempt an exhaustive description of GPU programming languages here, but refer the reader to a recent PhD thesis by Philipp Lucas, which contains an enlightening survey~\cite{Lucas08}. Lucas distinguishes carefully between languages (such as CG and HLSL) that aim to raise the level of abstraction at which graphics-oriented GPU programs are written, and those that attempt to abstract the entire GPU, and so must also provide a means to express the placing of programs on the GPU, feeding such programs with data, reading the results back to the CPU, and so on, as well as deciding to what extent the programmer should be involved
in stipulating those tasks.
In the first group of graphics-oriented languages, we include PyGPU and Vertigo.
PyGPU is a language for image processing embedded in Python~\cite{LUNDPAPER}. 
PyGPU uses the introspective abilities of Python and thus bypasses
the need to implement new loop structures and conditionals for the embedded 
language. In Python it is possible to access the bytecode of a function and 
from that extract information about loops and conditionals. 
Programs written in PyGPU can be compiled and run on a GPU. 
Vertigo is another embedded language by Conal Elliot~\cite{VERTIGO}. It
is a language for 3D graphics that targets the DirectX 8.1 shader model, and
can be used to describe geometry, shaders and to generate textures. 
%Each sublanguage is given formal semantics \cite{VERTIGO}. From programs 
%written in Vertigo, assembly language programs are generated for execution on a 
%GPU.

@@

\subsection{Lessons learned so far in the project}
Our first lesson has been the gradual realisation that a key aspect of a usable
GPU programming language that exposes details of the GPU architecture to the user is the means to express where and when data is placed in and
read from the memory hierarchy.
We are accustomed, from our earlier experience in hardware design, to describing
and generating networks of communicating components -- something like data-flow graphs. We are, however, unused to needing to express choices about the use
of the various levels in a memory hierarchy. We believe that we need to develop programming idioms and language support for this. It seems likely, too, that such idioms will not be quite as specific to GPU programming as other aspects of our embedded language development. How to deal with control of access to
a memory hierarchy in a parallel system seems to be a central problem that must be tackled if we are to develop better parallel programming methods in general.
A typical example of a generic approach to this problem is
the language Sequoia, which aims to provide programmers with a means
to express how the memory hierarchy is to be used, where a relatively abstract description of
the platform, viewed as a tree of processing nodes and memories, is a parameter~\cite{Sequoia}. Thus, programmers should write very generic code, which can be compiled for many different platforms. This kind of platform independence is not our aim here, and we would like to experiment with programming idioms for control of memory access for the particular case of a CPU plus some form of highly parallel co-processor that accelerates some computations.

>>>>

\subsection{Discussion} \label{sec:disc}
\subsubsection{Our influences}
As mentioned above, our earlier work on Lava has provided the inspiration for
the combinator-oriented or hardware-like style of programming that we are exploring in Obsidian.
On the other hand, the {\em implementation} of Obsidian has been much influenced by
{Pan}, an embedded language for image synthesis developed by 
Conal Elliot~\cite{PAN}. Because of the computational complexity of image generation, C 
code is generated. This C code can then be compiled by an optimising compiler. 
Many ideas from the paper 
``Compiling Embedded Languages'', describing the implementation of Pan have been 
used in the implementation of Obsidian~\cite{Elliott03:CompileDSEL-JFP}.

\subsubsection{Related work on GPU and GPGPU programming languages}
We cannot attempt an exhaustive description of GPU programming languages here, but refer the reader to a recent PhD thesis by Philipp Lucas, which contains an enlightening survey~\cite{Lucas08}. Lucas distinguishes carefully between languages (such as CG and HLSL) that aim to raise the level of abstraction at which graphics-oriented GPU programs are written, and those that attempt to abstract the entire GPU, and so must also provide a means to express the placing of programs on the GPU, feeding such programs with data, reading the results back to the CPU, and so on, as well as deciding to what extent the programmer should be involved
in stipulating those tasks.
In the first group of graphics-oriented languages, we include PyGPU and Vertigo.
PyGPU is a language for image processing embedded in Python~\cite{LUNDPAPER}. 
PyGPU uses the introspective abilities of Python and thus bypasses
the need to implement new loop structures and conditionals for the embedded 
language. In Python it is possible to access the bytecode of a function and 
from that extract information about loops and conditionals. 
Programs written in PyGPU can be compiled and run on a GPU. 
Vertigo is another embedded language by Conal Elliot~\cite{VERTIGO}. It
is a language for 3D graphics that targets the DirectX 8.1 shader model, and
can be used to describe geometry, shaders and to generate textures. 
%Each sublanguage is given formal semantics \cite{VERTIGO}. From programs 
%written in Vertigo, assembly language programs are generated for execution on a 
%GPU.

@@

\subsubsection{Lessons learned so far in the project}
Our first lesson has been the gradual realisation that a key aspect of a usable
GPU programming language that exposes details of the GPU architecture to the user is the means to express where and when data is placed in and
read from the memory hierarchy.
We are accustomed, from our earlier experience in hardware design, to describing
and generating networks of communicating components -- something like data-flow graphs. We are, however, unused to needing to express choices about the use
of the various levels in a memory hierarchy. We believe that we need to develop programming idioms and language support for this. It seems likely, too, that such idioms will not be quite as specific to GPU programming as other aspects of our embedded language development. How to deal with control of access to
a memory hierarchy in a parallel system seems to be a central problem that must be tackled if we are to develop better parallel programming methods in general.
A typical example of a generic approach to this problem is
the language Sequoia, which aims to provide programmers with a means
to express how the memory hierarchy is to be used, where a relatively abstract description of
the platform, viewed as a tree of processing nodes and memories, is a parameter~\cite{Sequoia}. Thus, programmers should write very generic code, which can be compiled for many different platforms. This kind of platform independence is not our aim here, and we would like to experiment with programming idioms for control of memory access for the particular case of a CPU plus some form of highly parallel co-processor that accelerates some computations.

### 2013-09-30 src2/papp/paperThesis.tex
apa

\section*{Abstract}
Obsidian is a domain specific language for data-parallel programming 
on graphics processors (GPUs). It is embedded in the functional 
programming language Haskell. The user writes code using constructs 
familiar from Haskell (like {\tt map} and {\tt reduce}), recursion 
and some specially designed combinators for combining GPU programs. 
NVIDIA CUDA code is generated from these high level descriptions, and 
passed to the {\tt nvcc} compiler~\cite{CUDAGuide2.0}. Currently, we 
consider only the generation of single kernels, and not their coordination.

@@

%-------------------------------------------------------------------------------
\section{Introduction}

@@

Obsidian is a domain specific language for general purpose programming on 
GPUs (GPGPU), capable of generating code for modern NVIDIA GPUs. 
In~\cite{JSMSKC_IFL08} we describe a version of Obsidian that uses a monadic 
interface. In the current version of Obsidian the monad has been replaced 
by another data structure, closely related to Arrows~\cite{Arrows}, representing GPU programs. Although the details are 
left out because of space limitations, the use of this GPU program representation 
can be seen in section~\ref{sec:GPUPrograms}. For a description of the implementation 
see~\cite{JSTECH} and for more general information on embedded language 
implementation see~\cite{Elliott03:CompileDSEL-JFP}.

@@

%-------------------------------------------------------------------------------
\subsection{NVIDIA GPUs and CUDA}

@@

% ------------------------------------------------------------------------------
\subsection{Aims of Obsidian}

@@

%-------------------------------------------------------------------------------
\section{Programming in Obsidian}

@@

%-------------------------------------------------------------------------------
\subsection{Array Language}
\label{sec:ArrayLanguage}
\FloatBarrier
Arrays in Obsidian do not, like in C, name an area of memory. Instead, an 
array is represented by an abstract data type, called {\tt Arr}, supporting 
the following operations: {\tt (!)} for indexing, {\tt len} 
returns the length of an array and {\tt mkArr} that given a function from 
indexes to elements and a length gives an array. For example using these 
operations reversing an array can be accomplished as follows:

@@

%Array programs like those described here make up the building blocks used
%to form larger GPU programs. How to create a GPU program from these is 
%shown in the following section.

@@

%\FloatBarrier
\subsection{GPU Programs}
\label{sec:GPUPrograms}

@@

\begin{small}
\begin{verbatim} 
data a :-> b = ... 
\end{verbatim}
\end{small} 
\noindent
%The details of {\tt a :-> b} are shown in section~\ref{sec:Implementation}. 
\noindent
Informally we can think of {\tt a\nolinebreak~:->\nolinebreak~b} as representing programs that operate 
as illustrated in figure~\ref{fig:program}. This figure 
shows a program that performs some computation using a number of threads 
followed by a barrier synchronisation, and so on. The contents of the 
boxes marked with {\em Pure} can be thought of as containing an array 
program.% such as those in section~\ref{sec:ArrayLanguage}.
The type constructor {\tt :->} is a variant of an arrow, which in turn is a generalisation of a monad~\cite{JSTECH}.
%
\begin{figure}
\begin{center}
\includegraphics[width=.60\linewidth]{./papp/pictures/prg_intuit.jpg}
\caption{A GPU program of type  {\tt a :-> b} can be thought 
of as pure computations interspersed by syncs.}
\label{fig:program}
\end{center}
\end{figure}
\noindent
One way to create a GPU program is by using the function {\tt pure} of type {\tt (a -> b) -> a :-> b}.
%\begin{small}
%\begin{verbatim}
%pure :: (a -> b) -> a :-> b
%\end{verbatim}
%\end{small}
For example the array language program {\tt fmap (+1)}, that increments 
every element of an array, can be lifted to a GPU program: 
%
\begin{small}
\begin{verbatim}
incr :: Arr IntE :-> Arr IntE 
incr = pure $ fmap (+1)
\end{verbatim}
\end{small} 
 \noindent
GPU programs can be executed on the GPU from
{\em GHCi} (the Haskell interpreter) using the function {\tt execute}:

\begin{small}
\begin{verbatim}
execute :: (Flatten a, Flatten b) => (Arr a :-> Arr b) -> [a] -> IO [b]
\end{verbatim}
\end{small}
\noindent
%The class {\tt Flatten} will be explained in detail in 
%section~\ref{sec:Implementation}, but 
Instances of {\tt Flatten} are all the types that can be stored in the 
GPU memory. Examples of types that are in {\tt Flatten} are {\tt IntE}, 
{\tt FloatE}, {\tt BoolE}. Arrays and pairs of things that are in Flatten 
are also instances of {\tt Flatten}.
Here, {\tt execute} is used in order to run an instance of the {\tt incr} 
program on the GPU:

@@

\section{Outline of Code Generation}
\label{sec:CodeGen}
\FloatBarrier

This section briefly reviews the code generation process using 
a hyphothetical program as example. The program under consideration 
is this:

@@

\section{Case Studies}
\label{sec:CaseStudies}

\FloatBarrier
This section describes a few slightly larger Obsidian programs. The main purpose of this section is to show 
how we want to write GPGPU programs using Obsidian. 
%How the performance 
%of the generated code can be improved is discussed in section~\ref{sec:FutureWork}  
%%% Move this later.  XXX

\subsection{Sorting}

@@

%-------------------------------------------------------------------------------
\subsection{Parallel Prefix}
\FloatBarrier

@@

This perfect division of labour is not obtainable with the current 
implementation of the {\tt How} argument to {\tt sync}. However, one 
division of the work that has experimentally been shown to perform well, see 
table in next section,  
is to let each thread $tid$ perform the work of $tid$ and $tid + 256$. This 
program is shown below:

@@

%-------------------------------------------------------------------------------
\FloatBarrier
\section{Future work}
\label{sec:FutureWork}

@@

In order to get to the quite efficient version of the {\tt sklansky} kernel 
in section~\ref{sec:CaseStudies}, the experimental {\tt How} argument to {\tt sync} was needed. 
This needs to be 
explored further. {\tt How} as it is today, any function of type
{\tt IndexE -> [IndexE]}, offers too much freedom and with that the risk of
introducing errors. We are searching for an elegant model for expressing how and what 
to compute, and for ways to integrate this with
the language. This is our biggest research challenge.

Section~\ref{sec:CaseStudies}, on case studies, showed that it is possible to
generate quite efficient code from Obsidian programs. The current version 
generates efficient code for very specific uses of the {\tt two} combinator, 
for example the kind used in the {\tt sklansky} parallel prefix network. More 
work needs to be done in order to find a good way to produce efficient code 
for a larger set of combinators. 
This may very well involve designing new combinators that do not need
to be nested as the current ones do.

@@

%-------------------------------------------------------------------------------
\section{Related work}

@@

\section{Discussion and conclusion}

Obsidian is work in progress and there are many loose ends to tie up and 
paths left to explore. In section~\ref{sec:CaseStudies}, the 
strengths of Obsidian show; it is possible to express quite complex algorithms
using short and elegant programs. The case studies also show that it is possible
to generate quite efficient code from these high level descriptions. However, 
this needs more work in order to more reliably produce efficient code and
to broaden the available set of combinators for combining GPU programs.

%Another benefit of a higher level language such as Obsidian compared to 
%CUDA is the ability to reuse code. There is an example of this too in the 
%case studies section where a single reduction program can be used to generate
%code to compute the maximum, minimum and sum very easily.

@@

%In Obsidian it is easy to describe an initial prototype solution to a problem,
%such as for example the {\tt mySum} kernels in section~\ref{sec:SyncPar}. 
%It requires no deep knowledge of the GPU architecture but works out of the 
%box. The prototype implementation can be tweaked into a more efficient 
%implementation by performing small changes to the Obsidian code that often 
%lead to quite large differences in the generated C code. An example of this 
%is {\tt mySum1} against {\tt mySum2} where adding a single {\tt sync} leads 
%to a radically different C program.

@@

%Obsidian is also an improvement over CUDA in the area of code reuse. The 
%{\tt reduce} function from section~\ref{sec:CaseStudies} is an example of this. 
%Obsidian is also compositional in another way than CUDA. Reusing kernels as
%building blocks in other kernels is in CUDA not realistic, while in Obsidian 
%it is the preferred way to write programs. Take as an example of this the 
%mergers and sorters in section~\ref{sec:CaseStudies}. In CUDA you would design 
%one kernel from scratch implementing the merger and sorter simultaneously.

@@

\section*{Acknowledgments}
\noindent
This research is funded by the Swedish Research Council.

>>>>

\subsection*{Abstract}
Obsidian is a domain specific language for data-parallel programming 
on graphics processors (GPUs). It is embedded in the functional 
programming language Haskell. The user writes code using constructs 
familiar from Haskell (like {\tt map} and {\tt reduce}), recursion 
and some specially designed combinators for combining GPU programs. 
NVIDIA CUDA code is generated from these high level descriptions, and 
passed to the {\tt nvcc} compiler~\cite{CUDAGuide2.0}. Currently, we 
consider only the generation of single kernels, and not their coordination.

@@

%-------------------------------------------------------------------------------
\subsection{Introduction}

@@

Obsidian is a domain specific language for general purpose programming on 
GPUs (GPGPU), capable of generating code for modern NVIDIA GPUs. 
In~\cite{JSMSKC_IFL08} we describe a version of Obsidian that uses a monadic 
interface. In the current version of Obsidian the monad has been replaced 
by another data structure, closely related to Arrows~\cite{Arrows}, representing GPU programs. Although the details are 
left out because of space limitations, the use of this GPU program representation 
can be seen in subsection~\ref{sec:GPUPrograms}. For a description of the implementation 
see~\cite{JSTECH} and for more general information on embedded language 
implementation see~\cite{Elliott03:CompileDSEL-JFP}.

@@

%-------------------------------------------------------------------------------
\subsubsection{NVIDIA GPUs and CUDA}

@@

% ------------------------------------------------------------------------------
\subsubsection{Aims of Obsidian}

@@

%-------------------------------------------------------------------------------
\subsection{Programming in Obsidian}

@@

%-------------------------------------------------------------------------------
\subsubsection{Array Language}
\label{sec:ArrayLanguage}
\FloatBarrier
Arrays in Obsidian do not, like in C, name an area of memory. Instead, an 
array is represented by an abstract data type, called {\tt Arr}, supporting 
the following operations: {\tt (!)} for indexing, {\tt len} 
returns the length of an array and {\tt mkArr} that given a function from 
indexes to elements and a length gives an array. For example using these 
operations reversing an array can be accomplished as follows:

@@

%Array programs like those described here make up the building blocks used
%to form larger GPU programs. How to create a GPU program from these is 
%shown in the following subsection.

@@

%\FloatBarrier
\subsubsection{GPU Programs}
\label{sec:GPUPrograms}

@@

\begin{small}
\begin{verbatim} 
data a :-> b = ... 
\end{verbatim}
\end{small} 
\noindent
%The details of {\tt a :-> b} are shown in subsection~\ref{sec:Implementation}. 
\noindent
Informally we can think of {\tt a\nolinebreak~:->\nolinebreak~b} as representing programs that operate 
as illustrated in figure~\ref{fig:program}. This figure 
shows a program that performs some computation using a number of threads 
followed by a barrier synchronisation, and so on. The contents of the 
boxes marked with {\em Pure} can be thought of as containing an array 
program.% such as those in subsection~\ref{sec:ArrayLanguage}.
The type constructor {\tt :->} is a variant of an arrow, which in turn is a generalisation of a monad~\cite{JSTECH}.
%
\begin{figure}
\begin{center}
\includegraphics[width=.60\linewidth]{./papp/pictures/prg_intuit.jpg}
\caption{A GPU program of type  {\tt a :-> b} can be thought 
of as pure computations interspersed by syncs.}
\label{fig:program}
\end{center}
\end{figure}
\noindent
One way to create a GPU program is by using the function {\tt pure} of type {\tt (a -> b) -> a :-> b}.
%\begin{small}
%\begin{verbatim}
%pure :: (a -> b) -> a :-> b
%\end{verbatim}
%\end{small}
For example the array language program {\tt fmap (+1)}, that increments 
every element of an array, can be lifted to a GPU program: 
%
\begin{small}
\begin{verbatim}
incr :: Arr IntE :-> Arr IntE 
incr = pure $ fmap (+1)
\end{verbatim}
\end{small} 
 \noindent
GPU programs can be executed on the GPU from
{\em GHCi} (the Haskell interpreter) using the function {\tt execute}:

\begin{small}
\begin{verbatim}
execute :: (Flatten a, Flatten b) => (Arr a :-> Arr b) -> [a] -> IO [b]
\end{verbatim}
\end{small}
\noindent
%The class {\tt Flatten} will be explained in detail in 
%subsection~\ref{sec:Implementation}, but 
Instances of {\tt Flatten} are all the types that can be stored in the 
GPU memory. Examples of types that are in {\tt Flatten} are {\tt IntE}, 
{\tt FloatE}, {\tt BoolE}. Arrays and pairs of things that are in Flatten 
are also instances of {\tt Flatten}.
Here, {\tt execute} is used in order to run an instance of the {\tt incr} 
program on the GPU:

@@

\subsection{Outline of Code Generation}
\label{sec:CodeGen}
\FloatBarrier

This subsection briefly reviews the code generation process using 
a hyphothetical program as example. The program under consideration 
is this:

@@

\subsection{Case Studies}
\label{sec:CaseStudies}

\FloatBarrier
This subsection describes a few slightly larger Obsidian programs. The main purpose of this subsection is to show 
how we want to write GPGPU programs using Obsidian. 
%How the performance 
%of the generated code can be improved is discussed in subsection~\ref{sec:FutureWork}  
%%% Move this later.  XXX

\subsubsection{Sorting}

@@

%-------------------------------------------------------------------------------
\subsubsection{Parallel Prefix}
\FloatBarrier

@@

This perfect division of labour is not obtainable with the current 
implementation of the {\tt How} argument to {\tt sync}. However, one 
division of the work that has experimentally been shown to perform well, see 
table in next subsection,  
is to let each thread $tid$ perform the work of $tid$ and $tid + 256$. This 
program is shown below:

@@

%-------------------------------------------------------------------------------
\FloatBarrier
\subsection{Future work}
\label{sec:FutureWork}

@@

In order to get to the quite efficient version of the {\tt sklansky} kernel 
in subsection~\ref{sec:CaseStudies}, the experimental {\tt How} argument to {\tt sync} was needed. 
This needs to be 
explored further. {\tt How} as it is today, any function of type
{\tt IndexE -> [IndexE]}, offers too much freedom and with that the risk of
introducing errors. We are searching for an elegant model for expressing how and what 
to compute, and for ways to integrate this with
the language. This is our biggest research challenge.

Subsection~\ref{sec:CaseStudies}, on case studies, showed that it is possible to
generate quite efficient code from Obsidian programs. The current version 
generates efficient code for very specific uses of the {\tt two} combinator, 
for example the kind used in the {\tt sklansky} parallel prefix network. More 
work needs to be done in order to find a good way to produce efficient code 
for a larger set of combinators. 
This may very well involve designing new combinators that do not need
to be nested as the current ones do.

@@

%-------------------------------------------------------------------------------
\subsection{Related work}

@@

\subsection{Discussion and conclusion}

Obsidian is work in progress and there are many loose ends to tie up and 
paths left to explore. In subsection~\ref{sec:CaseStudies}, the 
strengths of Obsidian show; it is possible to express quite complex algorithms
using short and elegant programs. The case studies also show that it is possible
to generate quite efficient code from these high level descriptions. However, 
this needs more work in order to more reliably produce efficient code and
to broaden the available set of combinators for combining GPU programs.

%Another benefit of a higher level language such as Obsidian compared to 
%CUDA is the ability to reuse code. There is an example of this too in the 
%case studies subsection where a single reduction program can be used to generate
%code to compute the maximum, minimum and sum very easily.

@@

%In Obsidian it is easy to describe an initial prototype solution to a problem,
%such as for example the {\tt mySum} kernels in subsection~\ref{sec:SyncPar}. 
%It requires no deep knowledge of the GPU architecture but works out of the 
%box. The prototype implementation can be tweaked into a more efficient 
%implementation by performing small changes to the Obsidian code that often 
%lead to quite large differences in the generated C code. An example of this 
%is {\tt mySum1} against {\tt mySum2} where adding a single {\tt sync} leads 
%to a radically different C program.

@@

%Obsidian is also an improvement over CUDA in the area of code reuse. The 
%{\tt reduce} function from subsection~\ref{sec:CaseStudies} is an example of this. 
%Obsidian is also compositional in another way than CUDA. Reusing kernels as
%building blocks in other kernels is in CUDA not realistic, while in Obsidian 
%it is the preferred way to write programs. Take as an example of this the 
%mergers and sorters in subsection~\ref{sec:CaseStudies}. In CUDA you would design 
%one kernel from scratch implementing the merger and sorter simultaneously.

@@

\subsection*{Acknowledgments}
\noindent
This research is funded by the Swedish Research Council.

### 2013-09-30 src2/thesis.tex
apa

\chapter{Obsidian: A Domain Specific Embedded Language for Parallel Programming of Graphics Processors}
%\addcontentsline{toc}{chapter}{Obsidian: A Domain Specific Embedded Language for
%Parallel Programming of Graphics Processors}

@@

\chapter{Obsidian: GPU Computing Using Haskell}
%\addcontentsline{toc}{chapter}{Obsidian: GPU Computing Using Haskell}

@@

\chapter{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
%\addcontentsline{toc}{chapter}{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}

@@

\chapter{Counting and Occurrence Sort for GPUs using an Embedded Language}
%\addcontentsline{toc}{chapter}{Counting and Occurrence Sort for GPUs using an Embedded Language}

@@

\chapter{Simple and Compositional Reification of Monadic Embedded Languages}
%\addcontentsline{toc}{chapter}{Simple and Compositional Reification of Monadic Embedded Languages}

@@

\chapter{Parallel Programming in Haskell Almost for Free}
%\addcontentsline{toc}{chapter}{Parallel Programming in Haskell Almost for Free}

>>>>

\section{Obsidian: A Domain Specific Embedded Language for Parallel Programming of Graphics Processors}
%\addcontentsline{toc}{chapter}{Obsidian: A Domain Specific Embedded Language for
%Parallel Programming of Graphics Processors}

@@

\section{Obsidian: GPU Computing Using Haskell}
%\addcontentsline{toc}{chapter}{Obsidian: GPU Computing Using Haskell}

@@

\section{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
%\addcontentsline{toc}{chapter}{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}

@@

\section{Counting and Occurrence Sort for GPUs using an Embedded Language}
%\addcontentsline{toc}{chapter}{Counting and Occurrence Sort for GPUs using an Embedded Language}

@@

\section{Simple and Compositional Reification of Monadic Embedded Languages}
%\addcontentsline{toc}{chapter}{Simple and Compositional Reification of Monadic Embedded Languages}

@@

\section{Parallel Programming in Haskell Almost for Free}
%\addcontentsline{toc}{chapter}{Parallel Programming in Haskell Almost for Free}
